### 2025-05-20 writing/manuscript.tex
first draft -introduction

\section{Introduction}

>>>>

\pagebreak
\section{Introduction}
Equally qualified individuals in the labor market may face very different outcomes depending on their socioeconomic status (SES). A key driver of inequality according to  sociologists is due to differences in social capital (e.g., Bourdieu). A lack of social capital means a lack of access to individuals with influential (higher paid) jobs and job opportunities. In economic terms, it implies having worse outcomes when using one's network to find jobs conditional on the capacity on leveraging one's social network.\footnote{\textcolor{red}{fill in citaiton sociology reading from slides}}.

Referral hiring, the formal or informal process where firms ask workers to recommend qualified candidates for job opportunities, is a vivid labor market application where the role of differences in social capital becomes evident. As referrals must orginate from the networks of referrers, the composition of referrer networks becomes a crucial channel that may propagate inequality: Similar individuals across socio-demographic characteristics tend to form connections at higher rates \citep{mcpherson2001birds}, making across SES (low-to-high) connections less likely than same-SES connections \citep{chetty2022social_a}.\footnote{Chetty and colleagues suggests the share of high-SES connections among low-SES networks as the definition of social capital, as it correlates so strongly with labor market income.} This implies referrals will reflect differences in network structures even in the absence of biases in the referral procedure, i.e., referring at random from one's network according to some productivity criteria.

Yet, in case of gender, evidence shows referrals may be biased under substantial pay-for-performance incentives beyond what is attribuable to differences in network compositions \citep{beaman_job_2018,hederos2025gender}. A similar bias against low-SES may further exacerbate outcomes of low-SES individuals in the labor market: If job information are in the hands of a select few high-SES which low-SES have already limited network access to (social capital hypothesis), and high-SES referrers are biased against low-SES, referring other high-SES at higher rates than their network composition, we can be confident that referral hiring disadvantages low-SES. The empirical question we answer is whether there is a bias against low-SES once we account for the network SES compositon in a controlled setting.

In this study, we focus on the role of SES in referrals by investigating whether individuals who are asked to refer someone tend to refer a candidate of their own SES. We also explore potential mechanisms behind referral patterns under different incentives. To this end, we conducted a lab-in-the-field experiment with 734 participants in a Colombian university where different SES groups mix together. Participants were instructed to refer a qualified student for tasks similar to the Math and Reading parts of the national university entry exam (akin to SAT). To incentivize participants to refer qualified candidates, we set earnings dependent on referred candidates' actual university entry exam scores.

Referral hiring in the labor market range form formal referral program at firm asking employees to refer to simply passing on job opportunities between network members \citep{topa2019social}. As participants in our study are students at the university referring based on exam scores, we abstract away from formal referral programs with defined job openings. Our setting instead resembles situations where contacts share opportunities with each other. This also eliminates reputational concerns as there is no firm, which play a largers role in the labor market compared to our setting. At the same time, national university entry exam scores are objective, widely accepted measures of ability, and referrers in our setting possess accurate information about these signals.
